---
title: "Network Analysis - Exercise 2"
output: 
  pdf_document: default
  github_document: default
  md_document: default
date: "2024-03-23"
---

## 1. Import Libraries

```{r Setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidygraph)
library(ggraph)
library(readr)
library(dplyr)
library(igraph)
```

## 2. Import Data
```{r Data Import}

data <- read.csv("/Users/chien/Library/CloudStorage/OneDrive-McGillUniversity/5d_Network_Analysis/Exercise2/Fakebook_bus_dataset.csv")

```

## 3. Calculate Centralilty

```{r Centrality Calculation}

# to take the first two columns, and graph is directed
g <- graph_from_data_frame(data[,1:2], directed=TRUE) 

# Calculate degree centrality
degree_centrality <- degree(g)

# Calculate closeness centrality
closeness_centrality <- closeness(g, normalized=TRUE) # normalized=TRUE scales the values between 0 and 1

# Calculate betweenness centrality
betweenness_centrality <- betweenness(g, normalized=TRUE) # normalized=TRUE scales the values between 0 and 1

# view centrality values
print("Degree Centrality:")
print(degree_centrality[1:4])

print("Closeness Centrality:")
print(closeness_centrality[1:4])

print("Betweeness Centrality:")
print(betweenness_centrality[1:4])

```

## 4. Possible Consequences

The analysis of centrality measures for seats within the bus network, particularly focusing on Node B, reveals significant implications for interpersonal dynamics and information flow among the passengers, presumably employees. Node B, distinguished by the highest degree centrality of 8, facilitates unparalleled direct access to other individuals. This position, therefore, affords maximum exposure to the diverse array of passengers, potentially enhancing one's understanding of the organizational culture and employee interrelations swiftly. Nonetheless, the predominance of Node B comes with potential drawbacks. The immediate and extensive visibility afforded by this central seat might not always contribute positively to the development of trust and influence. Occupying such a pivotal position could inadvertently project an image of overassertiveness or presumptuousness, which might not be favorably received by all members of the organization.

Furthermore, the closeness centrality measures, particularly pronounced for Nodes B and C, highlight the efficiency with which occupants of these seats can reach out to or be reached by others within the network. This characteristic is particularly beneficial for quick integration into the social fabric of the organization and expedited acquisition of situational awareness regarding the company's dynamics. However, the strategic advantage of such positions might encounter limitations during periods of high communicative demand, where these nodes might become inadvertent chokepoints or subjects of overwhelming attention.

The betweenness centrality of Node B, marked at approximately 0.134, underscores its indispensable role in mediating the flow of dialogue among the employees. This unique vantage point could enable an individual to subtly influence or gauge the pulse of information exchange across the network. Yet, the proximity to potentially sensitive or confidential exchanges may place the occupant of this seat in awkward situations, where their presence might deter open communication, thereby inadvertently impeding the natural flow of information.

In summary, while the central position represented by Node B in the network graph offers numerous advantages for rapid social integration and influence within the organizational context, it also imposes certain responsibilities and potential social dilemmas. The choice of this seat could thus be seen as a double-edged sword, offering significant opportunities for engagement and influence, albeit coupled with challenges in navigating the complexities of workplace dynamics and interpersonal communication. 

## 5. Plot Network graph with labels and centrality values

``` {r Network Analysis}
# adjust node_size for visualisation
degree_centrality_norm <- degree_centrality / max(degree_centrality)
node_sizes <- degree_centrality_norm * 50 

# Prepare labels with degree centrality
node_labels <- paste(V(g)$name, "\n(", round(degree_centrality),
                     round(closeness_centrality, 3), round(betweenness_centrality, 3),")", sep=" ")

# Plot Graph
par(mar=c(1, 1, 1, 1))

# Plot the network
plot(g,
     layout = layout_with_kk(g),
     vertex.size=node_sizes, # Change this for different centrality measures
     vertex.label=node_labels, # Assuming the nodes have names
     vertex.label.cex=0.8, # Adjust label size
     edge.arrow.size=0.3,
     vertex.label.dist = 0.5,
     main="Network Graph",
     vertex.color="skyblue")

legend("bottomright", # Position of the legend. Other options: "topleft", "bottomleft", "topright"
       title="Centrality Measures", # Title of the legend
       cex=0.8, # Size of the legend text
       bty="n", # No box around the legend
       legend=c("Degree Centrality Closeness Centrality Betweenness Centrality"), # Legend text
       ) # Margin around the legend
```

